# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*



```{r}
date()
read.csv("data/Learning2014_SUBSET.csv", sep = ",", header = TRUE)
sub_learning2014$gender <- as.factor(sub_learning2014$gender)
str(sub_learning2014)
dim(sub_learning2014)

```
The data id based on the survey of approaches to learning, which was implemented on a statistics course in fall 2014. The data includes 166 subjects and 7 variables: gender, age, attitude towards statistics (on a scale of 1-5), learning strategies (variables 'deep', 'stra' and 'surf', on scales of 1-5), and exam points 

Next, I will explore the variables and their correlations:

```{r}
summary (sub_learning2014$gender) 
summary (sub_learning2014$Age)
summary (sub_learning2014$attitude)
summary (sub_learning2014$deep) 
summary (sub_learning2014$stra)
summary (sub_learning2014$surf)
summary (sub_learning2014$Points)

boxplot(sub_learning2014$Age)
boxplot (sub_learning2014$attitude)
boxplot (sub_learning2014$deep) 
boxplot (sub_learning2014$stra)
boxplot (sub_learning2014$surf)
boxplot (sub_learning2014$Points)

pairs(sub_learning2014[-1], col = sub_learning2014$gender)

```
The distribution of Age is skewed towards the lower end of the scale with 75 % of the observations falling between (17,27).
The distribution of attitude is normal, not skewed, and seemes to have no outliers.
The distribution of deep is quite normal, and has two potential outliers in the lower end of the distribution.
The distribution of stra looks normal and seems to have no outliers.
The distribution of surf also looks normal and has one potential outlier in the upper end of the distribution.
The distribution of Points is pretty normal too. The observations seem to focus a little more on the upper side of the distribution.







